Weighted Model Counting with Conditional Weights

### 2020-08-01 paper/paper.tex
Make some things clearer

\begin{itemize}
\item The Main Narrative
  \begin{enumerate}
  \item When weights are defined on literals, the measure on the free BA is
    fully independent.
  \item This means that the BA itself must be larger (i.e., have additional
    `meaningless' literals) to turn any probability distribution into an
    independent one.
  \item We show how we can define conditional weights on literals, allowing us
    to encode any probability distribution into a Boolean algebra that's not
    necessarily independent and thus can be smaller.
  \item We demonstrate a specific example of this by presenting a new way to
    encode Bayesian networks into instances of WMC and adapting a WMC algorithm
    (ADDMC) to run on the new format.
  \item We show that this results in significantly faster inference.
  \item We show that our encoding results in asymptotically fewer literals and
    fewer ADDs.
  \end{enumerate}
\item Introduce
  \begin{itemize}
  \item WMC \cite{DBLP:conf/aaai/SangBK05}
  \item ADDMC \cite{DBLP:conf/aaai/DudekPV20} and ADDs
    \cite{DBLP:journals/fmsd/BaharFGHMPS97}
  \item A Bayesian network is a directed acyclic graph with random variables as
    vertices that defines a probability distribution over them. The full
    probability distribution can be expressed as the product of the conditional
    probability distributions of all random variables. For discrete Bayesian
    networks (and we only consider discrete networks in this paper), each such
    distribution can be represented by a table known as a conditional
    probability table (CPT).
  \end{itemize}
\item[F] What are the main claims, what are the main takeaways, intuitive [???]
  of theorems to follow.
\item Our work is, in spirit, similar to... using ROBDDs for Inference in
  Bayesian Networks with Troubleshooting as an Example
  \cite{DBLP:conf/uai/NielsenWJK00}: this is an interesting approach where
  deterministic parts of a BN are expressed as Boolean functions, i.e.,
  extracting the logical from the probabilistic. Maybe I can make a case that my
  encoding (the textual version) kind of does the same, even though the
  algorithm doesn't.
\end{itemize}

@@

\begin{table}
  \centering
  \caption{A comparison of Boolean-algebraic (BA) and set-theoretic (ST)
    concepts for $2^X$ for some set $X$}
  \label{tbl:notation}
  \begin{tabular}{lccl}
    \toprule
    BA name & BA symbol & ST symbol & ST name \\
    \midrule
    bottom & $\bot$ & $\emptyset$ & empty set \\
    top & $\top$ & $X$ & \\
    meet, and & $\land$ & $\cap$ & intersection \\
    join, or & $\lor$ & $\cup$ & union \\
    complement, not & $\neg$ & $^c$ & complement \\
            & $\le$ & $\subseteq$ & subset relation, set inclusion \\
    atom & & & singleton, unit set \\
    \bottomrule
  \end{tabular}
\end{table}

@@

\todo[inline]{Completely rewrite this section. It needs to be very clear and
  introduce all relevant terminology.}
\todo[inline]{Would it be more accurate to replace the word `model' with
  `interpretation' in most situations?}

For any set $X$, we can interpret $2^X$ as either a power set or a Boolean
algebra---see \cref{tbl:notation} for a summary of the differences in
terminology and notation.

Let $\mathcal{L}$ be a propositional logic with atoms $U = \{ a, b \}$. See
\cref{tbl:notation_example} for an overview of the models and formulas in
$\mathcal{L}$ as well as their set-theoretic and Boolean-algebraic
representations. Note the differences in the meaning of the word `atom' between
the logical and the Boolean-theoretic interpretations. In the Boolean algebra
$2^{2^U}$, an atom is a set with a single element which corresponds to a model
of $\mathcal{L}$ (also known as an element of $2^U$); whereas an atom of
$\mathcal{L}$ is an atomic formula, i.e., an element of $U$. We will primarily
use the term `atom' to mean the former.

In the rest of the paper, for any set $U$, we will use set-theoretic notation
for $2^U$ and Boolean-algebraic notation for $2^{2^U}$, except for (Boolean)
atoms in $2^{2^U}$ that are denoted as $\{x\}$ for some model $x \in 2^U$.

@@

We build on the definitions of multiplication and projection in the ADDMC paper
\cite{DBLP:conf/aaai/DudekPV20} and define more operations that can be used to
manipulate functions from Boolean algebras to non-negative reals. All of these
operations have efficient implementations in the CUDD \cite{somenzi1998cudd}
package for manipulating ADDs (among other things) that is used by ADDMC
{\cite{DBLP:conf/aaai/DudekPV20}.

@@

\subsection{Limitations of Factorable Measures}

@@

\begin{lemma} \label{lemma:before_theorem}
  For any measure $\mu\colon 2^{2^U} \to \mathbb{R}_{\ge 0}$ and elements $a, b
  \in 2^{2^U}$,
  \begin{equation} \label{eq:to_prove}
    \mu(a \land b) = \mu(a)\mu(b)
  \end{equation}
  if and only if
  \begin{equation} \label{eq:to_prove2}
    \mu(a \land b) \cdot \mu(\neg a \land \neg b) = \mu(a \land \neg b)
    \cdot \mu(\neg a \land b).
  \end{equation}
\end{lemma}
\begin{proof}
  First, note that $a = (a \land b) \lor (a \land \neg b)$ and $(a \land b)
  \land (a \land \neg b) = 0$, so, by properties of a measure,
  \begin{equation} \label{eq:temp}
    \mu(a) = \mu(a \land b) + \mu(a \land \neg b).
  \end{equation}
  Applying \cref{eq:temp} and the equivalent expression for $\mu(b)$ allows us
  to rewrite \cref{eq:to_prove} as
  \[
    \mu(a \land b) = [\mu(a \land b) + \mu(a \land \neg b)][\mu(a \land b) +
    \mu(\neg a \land b)]
  \]
  which is equivalent to
  \begin{equation} \label{eq:temp6}
    \mu(a \land b)[1 - \mu(a \land b) - \mu(a \land \neg b) - \mu(\neg a \land b)] = \mu(a \land \neg b)\mu(\neg a \land b).
  \end{equation}
  Since $a \land b$, $a \land \neg b$, $\neg a \land b$, $\neg a \land \neg b$
  are pairwise disjoint and their supremum is $\top$,
  \[
    \mu(a \land b) + \mu(a \land \neg b) + \mu(\neg a \land b) + \mu(\neg a
    \land \neg b) = 1,
  \]
  and this allows us to rewrite \cref{eq:temp6} into \cref{eq:to_prove2}. As all
  transformations are invertible, the two expressions are equivalent.
\end{proof}

\todo[inline]{The proof is only valid for probability measures!}

\begin{theorem}
  A measure $\mu\colon 2^{2^U} \to \mathbb{R}_{\ge 0}$ is factorable if and only
  if
  \begin{equation} \label{eq:wmccondition}
  \mu(u \land v) = \mu(u)\mu(v)
  \end{equation}
  for all distinct $u, v \in U \cup \{ \neg w \mid w \in U \}$ such that $u \ne
  \neg v$.
\end{theorem}
\begin{proof}
  ($\Leftarrow$) For each $x \in U$, let $\nu_x\colon 2^{\{x\}} \to
  \mathbb{R}_{\ge 0}$ be defined by $\nu_x(\{ x \}) = \mu(x)$ and
  $\nu_x(\emptyset) = \mu(\neg x)$. Let $\mu_\nu$ be the measure induced by $\nu
  = \prod_{x \in U} \nu_x$. We will show that $\mu = \mu_\nu$. First, note that
  $\mu_\nu(\bot) = 0 = \mu(\bot)$ by \cref{prop:measure} and the definition of a
  measure. Second, let $a = \bigwedge_{u \in U} a_u$ be an atom in $2^{2^U}$
  such that $a_u \in \{ u, \neg u \}$ for all $u \in U$. Then
  \[
    \mu_\nu(a) = \nu(\{ a_u \mid u \in U \}) = \prod_{u \in U} \nu_u(\{a_u\}) =
    \prod_{u \in U} \mu(a_u) = \mu\left(\bigwedge_{u \in U} a_u \right) = \mu(a)
  \]

  Finally, note that if $\mu_\nu$ and $\mu$ agree on all atoms, then they must
  also agree on all other non-zero elements of the Boolean algebra.

  ($\Rightarrow$) For the other direction, we are given a factored weight
  function $\nu = \prod_{x \in U} \nu_x$, and we want to show that its induced
  measure $\mu_\nu$ satisfies \cref{eq:wmccondition}. Let $k_u, k_v \in U \cup
  \{\neg w \mid w \in U \}$ be such that $k_u \in \{ u, \neg u \}$, $k_v \in \{
  v, \neg v \}$, and $u \ne v$. We then want to show that
  \begin{equation} \label{eq:to_prove3}
    \mu_\nu(k_u \land k_v) = \mu_\nu(k_u)\mu_\nu(k_v)
  \end{equation}
  which is equivalent to
  \begin{equation} \label{eq:to_prove4}
    \mu_\nu(k_u \land k_v) \cdot \mu_\nu(\neg k_u \land \neg k_v) = \mu_\nu(k_u \land \neg k_v) \cdot \mu_\nu(\neg k_u \land k_v)
  \end{equation}
  by \cref{lemma:before_theorem}. Then
  \begin{align*}
    \mu_\nu(k_u \land k_v) &= \sum_{\{a\} \le k_u \land k_v} \nu(a) = \sum_{\{a\} \le k_u \land k_v} \prod_{x \in U} \nu_x(a) \\
                        &= \sum_{\{a\} \le k_u \land k_v} \nu_u(a_u)\nu_v(a_v) \prod_{x \in U \setminus \{ u, v \}} \nu_x(a) = \sum_{\{a\} \le k_u \land k_v} \nu_u(k_u)\nu_v(k_v) \prod_{x \in U \setminus \{ u, v \}} \nu_x(a) \\
    &= \nu_u(k_u)\nu_v(k_v) \sum_{\{a\} \le k_u \land k_v} \prod_{x \in U \setminus \{ u, v \}} \nu_x(a) = \nu_u(k_u)\nu_v(k_v)c,
  \end{align*}
  where $c = \sum_{\{a\} \le k_u \land k_v} \prod_{x \in U \setminus \{ u, v \}}
  \nu_x(a)$, and is the same for $\mu_\nu(\neg k_u \land k_v)$, $\mu_\nu(k_u
  \land \neg k_v)$, and $\mu_\nu(\neg k_u \land \neg k_v)$ as well. But then
  \cref{eq:to_prove4} becomes
  \[
    \nu_u(k_u)\nu_v(k_v)\nu_u(\neg k_u)\nu_v(\neg k_v)c^2 = \nu_u(k_u)\nu_v(\neg
    k_v)\nu_u(\neg k_u)\nu_v(k_v)c^2
  \]
  which is trivially true.
\end{proof}

@@

\begin{theorem}
  Let $\phi\colon 2^U \to \mathbb{R}_{\ge 0}$ be the function generated by
  \cref{alg:encoding}. Then $(\exists_U(\phi \cdot [\lambda_{X=x}]))(\emptyset)
  = \Pr(X = x)$.
\end{theorem}
\begin{proof}
  Let $\mathcal{V} = \{ X, Y_1, \dots, Y_n \}$. Then
  \begin{align*}
    (\exists_U (\phi \cdot [\lambda_{X=x}]))(\emptyset) &= \sum_{T \in 2^U} (\phi \cdot [\lambda_{X=x}])(T) = \sum_{\lambda_{X=x} \in T \in 2^U} \phi(T) = \sum_{\lambda_{X=x} \in T \in 2^U} \left( \prod_{Y \in \mathcal{V}} \mathrm{CPT}_Y \right)(T) \\
    &= \sum_{(y_1, \dots, y_n) \in \prod_{i=1}^n \im Y_i} \Pr(X = x, Y_1 = y_1, \dots, Y_n = y_n) = \Pr(X = x)
  \end{align*}
  by the following arguments:
  \begin{itemize}
  \item the proof of Theorem~1 in the ADDMC paper \cite{DBLP:conf/aaai/DudekPV20};
  \item if $\lambda_{X=x} \not\in T \in 2^U$, then $(\phi \cdot
    [\lambda_{X=x}])(T) = \phi(T) \cdot [\lambda_{X=x}](T \cap \{
    \lambda_{X=x} \}) = \phi(T) \cdot 0 = 0$;
  \item \cref{lemma:full_distribution};
  \item marginalisation of a probability distribution.
  \end{itemize}
\end{proof}

@@

\todo[inline]{Describe how this is an intermediate step before we get the $\phi$
from the algorithm. Each CPT is treated as a clause.}

The Bayesian network in \cref{fig:example_bn} can be represented in a textual
format as
\[
  \begin{array}{l r r l l}
    \lambda\sb{T=l} &\lambda\sb{T=m} &\lambda\sb{T=h} & &0 \\
                    &-\lambda\sb{T=l} &-\lambda\sb{T=m} & &0 \\
                    &-\lambda\sb{T=l} &-\lambda\sb{T=h} & &0 \\
                    &-\lambda\sb{T=m} &-\lambda\sb{T=h} & &0 \\
    w &\lambda\sb{W=1} & &0.5 &0.5 \\
    w &\lambda\sb{F=1} &\lambda\sb{W=1} &0.6 &0.4 \\
    w &\lambda\sb{F=1} &-\lambda\sb{W=1} &0.1 &0.9 \\
    w &\lambda\sb{T=l} &\lambda\sb{W=1} &0.2 &1 \\
    w &\lambda\sb{T=m} &\lambda\sb{W=1} &0.4 &1 \\
    w &\lambda\sb{T=h} &\lambda\sb{W=1} &0.4 &1 \\
    w &\lambda\sb{T=l} &\lambda\sb{W=0} &0.6 &1 \\
    w &\lambda\sb{T=m} &\lambda\sb{W=0} &0.3 &1 \\
    w &\lambda\sb{T=h} &\lambda\sb{W=0} &0.1 &1
  \end{array}
\]
with each $\lambda$ replaced with a unique positive integer. This format is
based on the format used by the Cachet solver \cite{DBLP:conf/sat/SangBBKP04}
to encode WMC problems, which extends the DIMACS format for conjunctive normal
form (CNF) formulas with weight clauses. Subsequently, we extend it in two ways:
a single weight clause now supports an arbitrary number of literals, and each
weight clause has two probabilities instead of one (i.e., we no longer assume
that $\Pr(v) + \Pr(\neg v) = 1$ for all $v \in U$). In our use of the encoding,
it is always the case that either both probabilities sum to one, or the second
probability (i.e., the probability for the complement of the variable) is equal
to one.

@@

ADDMC constructs the \emph{Gaifman graph} \cite{gaifman1982local} of the input
CNF formula as an aid for the algorithm's heuristics. This graph has as vertices
the variables of the formula, and there is an edge between two variables $u$ and
$v$ if there is a clause in the formula that contains both $u$ and $v$. We
extend this definition to functions on Boolean algebras, i.e., the factors of
$\phi$. For any pair of distinct variables $u, v \in U$, we draw an edge between
them in the Gaifman graph if there is a function $A\colon 2^X \to
\mathbb{R}_{\ge 0}$ that is a factor of $\phi$ such that $u \in X$ and $v \in
X$. For instance, a factor such as $\mathrm{CPT}_X$ will enable edges between
all distinct pairs of variables in $\mathcal{E}^*(X)$.

\todo[inline]{This needs rewriting.}
ADDMC multiplies the final answer by $w(u)+w(-u)$ for every $u \in U$ that was
`simplified out' of the ADD for $\phi$ and so is not featured in the $\exists_U$
projection. In our format, for every such $u \in U$, we multiply the answer by
two. Each such $u \in U$ must satisfy two properties:
\begin{itemize}
\item all the probabilities/weights associated with $u$ are equal to $0.5$
  (otherwise the corresponding CPT would depend on $u$),
\item and all other CPTs are independent of $u$ (or they may have a trivial
  dependence, where the probability stays the same if $u$ is replaced with its
  complement).
\end{itemize}
Thus, the CPT that corresponds to $u$ still multiplies every model by $0.5$, but
the number of models being considered by the ADDMC is halved. To correct for
this, we need to perform this multiplication.

@@

\paragraph{Observations.}
\begin{itemize}
\item The order of the encodings from best to worst is exactly the opposite of
  that in the previous literature.
\item Our encoding is particularly promising on instances of Grid networks.
  They are set up so that the entire Bayesian network is relevant to the
  query.
\item However, \texttt{db20} struggles with instances from Mastermind and
  Random Blocks domains (but so does \texttt{sbk05}). We conjecture that this
  is so either because the particular structure of these problems confuse the
  heuristics used by ADDMC or because these instances allow for significant
  simplifications that are exploited by $\mathtt{d02}$ but not
  $\mathtt{db20}$.
\item Everything \texttt{d02} can do in \SI{1000}{\second}, \texttt{db20} can do
  in \SI{10}{\second} (find the exact number).
\end{itemize}

\subsection{Explaining the Performance Benefits}

\todo[inline]{Maybe replace `upper bound' with `worst-case'.}

@@

Let $n = |\mathcal{V}|$ be the number of vertices in the Bayesian network, $d =
\max_{X \in \mathcal{V}} |\mathrm{pa}(X)|$ the maximum in-degree (i.e., number
of parents), and $v = \max_{X \in \mathcal{V}} |\im X|$ the maximum number of
values per variable. Then... \cref{tbl:asymptotes}. However, note that these are
upper bounds and most encodings (including \texttt{db20}) can be smaller in
certain situations (e.g., with binary random variables or when a CPT has
repeating probabilities).

We equate clauses and ADDs (more specifically, factors of the function $\phi$
from \cref{alg:encoding}) here because ADDMC interprets each clause of any WMC
encoding as a multiplicative factor of the ADD that represents the entire WMC
instance \cite{DBLP:conf/aaai/DudekPV20}. For literal-weight encodings, each
weight is also a factor, but that has no effect on the asymptotic bounds in the
table.

\section{Conclusion and Future Work}

>>>>

\begin{itemize}
\item The main narrative
  \begin{enumerate}
  \item Putting weights on literals is restrictive.
  \item To overcome this, it is common practice to add more literals.
  \item This is indeed always possible but much slower.
  \item We show how we can define conditional weights on literals, allowing us
    to encode any measure.
  \item We demonstrate a specific example of this by presenting a new way to
    encode Bayesian networks into instances of WMC and adapting a WMC algorithm
    (ADDMC) to run on the new format.
  \item We show that this results in significantly faster inference.
  \item We show that our encoding results in asymptotically fewer literals and
    fewer ADDs.
  \end{enumerate}
\item Introduce
  \begin{itemize}
  \item WMC \cite{DBLP:conf/aaai/SangBK05}
  \item ADDMC \cite{DBLP:conf/aaai/DudekPV20} and ADDs
    \cite{DBLP:journals/fmsd/BaharFGHMPS97}
  \item A Bayesian network is a directed acyclic graph with random variables as
    vertices that defines a probability distribution over them. The full
    probability distribution can be expressed as the product of the conditional
    probability distributions of all random variables. For discrete Bayesian
    networks (and we only consider discrete networks in this paper), each such
    distribution can be represented by a table known as a conditional
    probability table (CPT).
  \end{itemize}
\item[F] What are the main claims, what are the main takeaways, intuitive [???]
  of theorems to follow.
\item Our work is, in spirit, similar to... using ROBDDs for Inference in
  Bayesian Networks with Troubleshooting as an Example
  \cite{DBLP:conf/uai/NielsenWJK00}: this is an interesting approach where
  deterministic parts of a BN are expressed as Boolean functions, i.e.,
  extracting the logical from the probabilistic. Maybe I can make a case that my
  encoding (the textual version) kind of does the same, even though the
  algorithm doesn't.
\end{itemize}

@@

\begin{table}
  \centering
  \caption{A comparison of Boolean-algebraic (BA) and set-theoretic (ST)
    concepts for $2^U$ for any set $U$}
  \label{tbl:notation}
  \begin{tabular}{lccl}
    \toprule
    BA name & BA symbol & ST symbol & ST name \\
    \midrule
    bottom & $\bot$ & $\emptyset$ & empty set \\
    top & $\top$ & $U$ & \\
    meet, and & $\land$ & $\cap$ & intersection \\
    join, or & $\lor$ & $\cup$ & union \\
    complement, not & $\neg$ & $^c$ & complement \\
            & $\le$ & $\subseteq$ & subset relation, set inclusion \\
    atom & & & singleton, unit set \\
    \bottomrule
  \end{tabular}
\end{table}

@@

Let $\mathcal{L}$ be a propositional logic with atoms $a$ and $b$, and let $U =
\{ a, b \}$. Then $2^U$, the power set of $U$, is the set of all models of
$\mathcal{L}$, and $2^{2^U}$ is the set of all formulas. These sets can also be
represented as Boolean algebras---see \cref{tbl:notation} for an overview of
notational differences and \cref{tbl:notation_example} for examples of how
various elements can be represented in both notations. Most importantly, note
that the word \emph{atom} has completely different meanings in logic and in
Boolean algebras. An atom in $\mathcal{L}$ is an atomic formula, i.e., an
element of $U$, whereas an atom in a Boolean algebra is (in set-theoretic terms)
a singleton set. For instance, an atom in $2^{2^U}$ corresponds to a model of
$\mathcal{L}$, i.e., an element of $2^U$. Unless referring specifically to a
logic, we will use the algebraic definition of an atom while referring to
logical atoms as \emph{variables}. In the rest of the paper, for any set $U$, we
will use set-theoretic notation for $2^U$ and Boolean-algebraic notation for
$2^{2^U}$, except for (Boolean) atoms in $2^{2^U}$ that are denoted as $\{x\}$
for some model $x \in 2^U$.

@@

We build on the definitions of multiplication and projection in the ADDMC paper
\cite{DBLP:conf/aaai/DudekPV20} and define more operations that can be used to
manipulate functions from Boolean algebras to non-negative real numbers. All of
these operations have efficient implementations in the CUDD
\cite{somenzi1998cudd} package for manipulating ADDs (among other things) that
is used by ADDMC {\cite{DBLP:conf/aaai/DudekPV20}.

@@

\subsection{Not All Measures Are Factorable}

@@

\todo[inline]{Add the proof that one can always go from any measure to a
  factorable measure by adding more literals and say: Not all measures are
  factorable but it is common practice to add more literals to make the measure
  factorable. We show that it's always possible and yet it can be faster not to
  do it.}

@@

\begin{theorem}
  Let $\phi\colon 2^U \to \mathbb{R}_{\ge 0}$ be the function generated by
  \cref{alg:encoding}. Then $(\exists_U(\phi \cdot [\lambda_{X=x}]))(\emptyset)
  = \Pr(X = x)$.
\end{theorem}
\begin{proof}
  Let $\mathcal{V} = \{ X, Y_1, \dots, Y_n \}$. Then
  \begin{align*}
    (\exists_U (\phi \cdot [\lambda_{X=x}]))(\emptyset) &= \sum_{T \in 2^U} (\phi \cdot [\lambda_{X=x}])(T) = \sum_{\lambda_{X=x} \in T \in 2^U} \phi(T) = \sum_{\lambda_{X=x} \in T \in 2^U} \left( \prod_{Y \in \mathcal{V}} \mathrm{CPT}_Y \right)(T) \\
    &= \sum_{(y_1, \dots, y_n) \in \prod_{i=1}^n \im Y_i} \Pr(X = x, Y_1 = y_1, \dots, Y_n = y_n) = \Pr(X = x)
  \end{align*}
  by the following arguments:
  \begin{itemize}
  \item the proof of Theorem~1 in the ADDMC paper \cite{DBLP:conf/aaai/DudekPV20};
  \item if $\lambda_{X=x} \not\in T \in 2^U$, then $(\phi \cdot
    [\lambda_{X=x}])(T) = \phi(T) \cdot [\lambda_{X=x}](T \cap \{
    \lambda_{X=x} \}) = \phi(T) \cdot 0 = 0$;
  \item \cref{lemma:full_distribution};
  \item marginalisation of a probability distribution.
  \end{itemize}
\end{proof}
\todo[inline]{Need to make the proof nicer.}

@@

\Cref{alg:encoding} encodes a Bayesian network into a function on a Boolean
algebra, but how does it relate to the standard interpretation of a WMC encoding
as a formula in conjunctive normal form (CNF) together with a collection of
weights? The factors of $\phi$ that restrict the values of indicator variables
for non-binary random variables are already expressed as a product of sums of
0/1-valued functions, i.e., a kind of CNF. Disregarding these functions, each
conditional weight function $\mathrm{CPT}_X$ is represented by a sum with a term
for every subset of $\mathcal{E}^*(X)$. To encode these terms, we introduce
\emph{extended weight clauses} to the WMC format used by Cachet
\cite{DBLP:conf/sat/SangBBKP04}. For instance, here is a representation of the
Bayesian network from \cref{fig:example_bn}:
\[
  \begin{array}{lrrll}
    \lambda\sb{T=l} &\lambda\sb{T=m} &\lambda\sb{T=h} & &0 \\
                    &-\lambda\sb{T=l} &-\lambda\sb{T=m} & &0 \\
                    &-\lambda\sb{T=l} &-\lambda\sb{T=h} & &0 \\
                    &-\lambda\sb{T=m} &-\lambda\sb{T=h} & &0 \\
    w &\lambda\sb{W=1} & &0.5 &0.5 \\
    w &\lambda\sb{F=1} &\lambda\sb{W=1} &0.6 &0.4 \\
    w &\lambda\sb{F=1} &-\lambda\sb{W=1} &0.1 &0.9 \\
    w &\lambda\sb{T=l} &\lambda\sb{W=1} &0.2 &1 \\
    w &\lambda\sb{T=m} &\lambda\sb{W=1} &0.4 &1 \\
    w &\lambda\sb{T=h} &\lambda\sb{W=1} &0.4 &1 \\
    w &\lambda\sb{T=l} &-\lambda\sb{W=1} &0.6 &1 \\
    w &\lambda\sb{T=m} &-\lambda\sb{W=1} &0.3 &1 \\
    w &\lambda\sb{T=h} &-\lambda\sb{W=1} &0.1 &1
  \end{array}
\]
where each indicator variable is eventually replaced with a unique positive
integer. Each line prefixed with a $w$ can be split into four parts: the `main'
variable (always not negated), conditions (possibly none), and two weights. For
example, the line
\[
  \begin{array}{lrrll}
    w &\lambda\sb{T=m} &-\lambda\sb{W=1} &0.3 &1
  \end{array}
\]
encodes the function $0.3[\lambda_{T=m}] \cdot \overline{[\lambda_{W=1}]} +
1\overline{[\lambda_{T=m}]} \cdot \overline{[\lambda_{W=1}]}$ and can be
interpreted as defining two conditional weights: $\nu(T = m \mid W = 0) = 0.3$,
and $\nu(T \ne m \mid W = 0) = 1$, the former of which corresponds to a row in
the CPT of $T$ while the latter is artificially added as part of the encoding.
In our encoding of Bayesian networks, it is always the case that, in each weight
clause, either both weights sum to one, or the second weight is equal to one.
Finally, note that (without any additional restrictions), the measure induced by
these weight functions is not probabilistic (i.e., $\mu(\top)$ may not be equal
to one).

@@

ADDMC constructs the \emph{Gaifman graph} \cite{gaifman1982local} of the input
CNF formula as an aid for the algorithm's heuristics. This graph has as vertices
the variables of the formula, and there is an edge between two variables $u$ and
$v$ if there is a clause in the formula that contains both $u$ and $v$. We
extend this definition to functions on Boolean algebras, i.e., the factors of
$\phi$. For any pair of distinct variables $u, v \in U$, we draw an edge between
them in the Gaifman graph if there is a function $\alpha\colon 2^X \to
\mathbb{R}_{\ge 0}$ that is a factor of $\phi$ such that $u \in X$ and $v \in
X$. For instance, a factor such as $\mathrm{CPT}_X$ will enable edges between
all distinct pairs of variables in $\mathcal{E}^*(X)$.

Even though the function $\phi$ produced by \cref{alg:encoding} is constructed
to have $2^U$ as its domain, sometimes the domain is effectively reduced to
$2^V$ for some $V \subset U$ by the ADD manipulation algorithms that optimise
the ADD representation of a function. For a simple example, consider $\alpha:
2^{\{a\}} \to \mathbb{R}_{\ge 0}$ defined as $\alpha(\{a\}) = \alpha(\emptyset)
= 0.5$. Then $\alpha$ can be reduced to $\alpha'\colon 2^{\emptyset} \to
\mathbb{R}_{\ge 0}$ defined as $\alpha'(\emptyset) = 0.5$. To compensate for
these reductions, for the original WMC format with a weight function $w\colon U
\cup \{ \neg u \mid u \in U \} \to \mathbb{R}_{\ge 0}$, ADDMC would multiply its
computed answer by $\prod_{u \in U \setminus V} w(u) + w(\neg u)$. With the new
WMC format, we instead multiply the answer by $2^{|U \setminus V|}$. Each
`excluded' variable $u \in U \setminus V$ satisfies two properties:
all weights associated with $u$ are equal to $0.5$ (otherwise the corresponding
CPT would depend on $u$, and $u$ would not be excluded), and all other CPTs are
independent of $u$ (or they may have a trivial dependence, where the probability
stays the same if $u$ is replaced with its complement).
Thus, the CPT that corresponds to $u$ still multiplies every model by $0.5$, but
the number of models being considered by the ADDMC is halved. To correct for
this, we multiply the final answer by two for every $u \in U \setminus V$.

@@

\paragraph{Observations.}
\begin{itemize}
\item The order of the encodings from best to worst is exactly the opposite of
  that in the previous literature.
\item Our encoding is particularly promising on instances of Grid networks.
  They are set up so that the entire Bayesian network is relevant to the
  query.
\item However, \texttt{db20} struggles with instances from Mastermind and
  Random Blocks domains (but so does \texttt{sbk05}). We conjecture that this
  is so either because the particular structure of these problems confuse the
  heuristics used by ADDMC or because these instances allow for significant
  simplifications that are exploited by $\mathtt{d02}$ but not
  $\mathtt{db20}$.
\item An observation from the cumulative plot: \texttt{db20} solves the same
  number of instances in \SI{6.374}{\second} as \texttt{d02} does in
  \SI{1000}{\second}.
\end{itemize}

@@

\todo[inline]{It could be confusing to use the word `variable' here. I'm using
  it as a synonym to `logical atom'.}

\paragraph{Explaining the performance benefits.} Let $n = |\mathcal{V}|$ be the
number of vertices in the Bayesian network, $d = \max_{X \in \mathcal{V}}
|\mathrm{pa}(X)|$ the maximum in-degree (i.e., number of parents), and $v =
\max_{X \in \mathcal{V}} |\im X|$ the maximum number of values per variable.
\Cref{tbl:asymptotes} shows how \texttt{db20} has both fewer variables and fewer
ADDs than any other encoding. However, note that these are upper bounds and most
encodings (including \texttt{db20}) can be smaller in certain situations (e.g.,
with binary random variables or when a CPT has repeating probabilities). We
equate clauses and ADDs (more specifically, factors of the function $\phi$ from
\cref{alg:encoding}) here because ADDMC interprets each clause of any WMC
encoding as a multiplicative factor of the ADD that represents the entire WMC
instance \cite{DBLP:conf/aaai/DudekPV20}. For literal-weight encodings, each
weight is also a factor, but that has no effect on the asymptotic bounds in the
table.

% While some ADDs may be larger as a result, experimental evidence still shows
% significant performance benefits.

\section{Conclusions and Future Work}
